### 2021-07-07 response.tex

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}

\usepackage{color}

\usepackage{xspace}

\usepackage{lmodern}
\usepackage{amssymb,amsmath}

\usepackage[pdfencoding=auto, psdextra]{hyperref}

\usepackage{natbib}
\bibliographystyle{chicago}

\newcommand{\eref}[1]{Eq.~(\ref{eq:#1})}
\newcommand{\fref}[1]{Fig.~\ref{fig:#1}}

\newcommand{\rR}{\mbox{$r$--$\cal R$}}
\newcommand{\RR}{\ensuremath{{\cal R}}}
\newcommand{\RRhat}{\ensuremath{{\hat \cal R}}}
\newcommand{\Rx}[1]{\ensuremath{{\cal R}_{#1}}} 
\newcommand{\Ro}{\ensuremath{{\mathcal R}_{0}}\xspace}
\newcommand{\Rs}{\Rx{\mathrm{s}}}
\newcommand{\Rpool}{\ensuremath{{\mathcal R}_{\textrm{\tiny{pool}}}}\xspace}
\newcommand{\Reff}{\Rx{\mathit{eff}}}
\newcommand{\Tc}{\ensuremath{C}}

\newcommand{\dd}[1]{\ensuremath{\, \mathrm{d}#1}}
\newcommand{\dtau}{\dd{\tau}}
\newcommand{\dx}{\dd{x}}
\newcommand{\dsigma}{\dd{\sigma}}

\newcommand{\rev}{\subsection*}
\newcommand{\revtext}{\textsf}
\setlength{\parskip}{\baselineskip}
\setlength{\parindent}{0em}

\newcommand{\comment}[3]{\textcolor{#1}{\textbf{[#2: }\textsl{#3}\textbf{]}}}
\newcommand{\jd}[1]{\comment{cyan}{JD}{#1}}
\newcommand{\swp}[1]{\comment{magenta}{SWP}{#1}}
\newcommand{\dc}[1]{\comment{blue}{DC}{#1}}
\newcommand{\jsw}[1]{\comment{green}{JSW}{#1}}
\newcommand{\hotcomment}[1]{\comment{red}{HOT}{#1}}

\newcommand{\psymp}{\ensuremath{p}} %% primary symptom time
\newcommand{\ssymp}{\ensuremath{s}} %% secondary symptom time
\newcommand{\pinf}{\ensuremath{\alpha_1}} %% primary infection time
\newcommand{\sinf}{\ensuremath{\alpha_2}} %% secondary infection time

\newcommand{\psize}{{\mathcal P}} %% primary cohort size
\newcommand{\ssize}{{\mathcal S}} %% secondary cohort size

\newcommand{\gtime}{\tau_{\rm g}} %% generation interval
\newcommand{\gdist}{g} %% generation-interval distribution
\newcommand{\idist}{\ell} %% incubation period distribution

\newcommand{\total}{{\mathcal T}} %% total number of serial intervals

\begin{document}

\noindent Dear Editor:

Thank you for the chance to revise and resubmit our manuscript. 
We have made major revisions to our manuscript to address the reviewers' comments.

\rev{Reviewer \#1}

\revtext{This is a review of ``Roles of generation-interval distributions in shaping relative epidemic strength, speed, and control of new SARS-CoV-2 variants'' by Park and others for Royal Society Interface.  This is the latest in an ongoing series of papers (Champredon 2015, Champredon 2018, Park 2019, Dushoff 2021, ...) in which the last author has considered aspects of incubation, growth, and reproduction in SIR epidemic theory.  In this manuscript, the authors use theory to anticipate the effects of COVID-19 variants where the distribution of infectivity evolves.}

\revtext{The manuscript is excellently written. The figures are clear.  All the mathematics seems satisfactory.  Unfortunately, the manuscript does not make its case for novel results.}

Thank you.

\revtext{The title is overly-broad as it suggests consideration of general forms of generation interval distributions (along the lines of Miller 2010 10.1016/j.jtbi.2009.08.007), when infect the authors study variation around the time of peak infectivity using a (very standard) $\gamma$-chain trick.  The study of generation time and $\gamma$ chains has a long history from the cited-but-unreviewed Svensson 2007, Wallinga 2007, to uncited Kenah 2008, A.  Lloyd 2001, D. Anderson 1980, Bailey 1964 (see Park 2019 and Champredon 2015).  In light of these and other works, the title and abstract fail to identify novel contributions to our understanding of epidemic theory.}

Thank you for the suggestions.

\revtext{Although the authors are aware of the history of research on this topic, they do little to review the relevant material, leaving this manuscript feeling like a superficial re-hashing of obvious material.  The introduction doesn't provide any serious review of the theory}

Done. We do not review gamma chain trick in depth because we're not doing it.

\revtext{Section 4 doesn't give references to prior work, and section 5 ignores almost all of the literature on the topic of estimating R from incidence data.}

Section 4 done. Section 5 done.

\revtext{Other comments:}

\revtext{Topic on which there is a considerable confusion in terminology in the literature.  In addition to "generation interval", terms "generation time", "transmission interval", "serial interval", and others have been used in the literature, all with related by different definitions.  This manuscripts definition "time between infection and transmission" implies a single transmission event, but Kenah pointed out that time-contraction appears when allowing multiple transmission events (see Champredon 2015).  A clearer definition up front will help readers.  Fortunately, the results don't depend on the specific definition.}

Done.

\revtext{The 'speed'-'strength' distinction from the previous paper is seductive, but perhaps leaning into general quantitative illiteracy.  When one pays attention to units -- "speed" L/T,  "Growth rate" 1/T while "strength" has dimensions of force, typically, but reproduction number is a dimensionless ratio of cases per case.  Beyond time-independent final-size relations (which are themselves not robust to effects like behavior change), several researchers have argued that growth rates generally more useful than reproduction numbers (giving equivalent threshold conditions, for example) and not as prone to mis-interpretation.  Other authors in evolutionary biology and economics have argued that the reproduction number's scientific value is best salvaged using the "discounted reproduction success" (McNamara, Houston, and Collins 2001), rather than fighting of an R-r straw-man.}

No.

\revtext{Equation 1 As Kermack and McKendrick introduced it, the compartmental model is a convenient *specialization* of the integral models-- see the nice re-analysis by Breda et al., 2012.}

Thank you for pointing this out. We now cite Breda et al. We acknowledge that ``This framework ... generalizes compartmental model, such as the SEIR model''

\revtext{Equation 4 assumes the wild-type and variant co-dominate the population.  This may not be the cases when multiple variants are co-circulating, particularly during replacement events.}

Done.

\revtext{p 4 l55.  Shape of the distribution matters -- heavy tails and failed central concentration.  This detail will escape the naive reader unless we poke them a little more with it.}

Reworded:

``For simplicity we neglect differences in the squared coefficient of variation and assume $\kappa_{\mathrm{wt}} = \kappa_{\mathrm{var}} = \kappa$.
While the exact shape of the distribution matters, we choose to focus on differences in the mean generation intervals to provide a qualitative understanding of how differences in generation-interval distributions between the wild type and the variant affects the relationship between relative strength and speed.
In general, we expect  differences in the squared coefficient of variation to have qualitatively similar effects as differences in the mean.
For example, a wider distribution allows for more early transmission, which in turn reduces epidemic strength given speed---this effect is qualitatively similar to that of a distribution with a shorter mean.''

\rev{Reviewer \#2}

\revtext{This is a well-written paper about the role of the generation interval distribution in the estimation of the relative strength (reproduction number) and speed
(epidemic growth rate) of a new variant of a disease. Using a renewal equation
framework, they consider inferring relative strength from relative speed and vice
versa as well as inferring relative strength from incidence data. While all of the
analysis in the paper is sound, I am concerned that the strong assumptions made
(mass-action epidemic, constant generation interval distribution, etc.) will limit
the reliability of the results in a practical application. However, they clearly
establish that assumptions about the generation interval distribution have important implications for the analysis of novel variants. While I am generally
skeptical of population-level analyses of infectious disease transmission based
on generation intervals, these are done frequently so this is a useful insight.
More detailed comments are below.}

Thank you. We have tried to make the limitation more clear.

\revtext{1. (page 3, first paragraph) It is important to acknowledge that the renewal equation approach to the relationship between the generation interval
distribution, R, and the growth rate r assumes a mass-action epidemic.
When we assume a fixed generation interval distribution, we must add the
assumption that the rate of depletion of susceptibles over time is negligible and the prevalence of infection is low. One of the most important
findings of Ref 10 is that the mean generation interval contracts when the
rate of depletion of susceptibles increases. While the renewal equation
does allow for the generation interval to change over time, the analyses
in Sections 3–5 assume fixed mean generation intervals. Because variants
are likely to spread when susceptibles are being depleted by both variants
(due to cross-protection) and when the prevalence of infection is not low, these implicit assumptions place significant limitations on the practical
implications of the analysis.}

This is a confusion between the intrinsic and realized. Also done.

\revtext{ (page 2, lines 55–60) It would help to clarify what is meant by “strengthlike” and “speed-like” in interventions. From the examples, it appears that
speed-like interventions work by identifying potential infections quickly
(effectively ending the infectious period early while not necessarily altering
the hazard of infectious contact) and strength-like interventions work by
reducing the hazard of infectious contact (while not necessarily altering
the duration of infectiousness). A good explanation of these ideas is on
lines 228–231 on page 10.}

Done.

\revtext{(page 5, lines 126–128 and 137) It would help to point out that you assume
rwt < rvar when you introduce the five scenarios.}

Done.

\revtext{(page 9, Figure 3) I would order the rows in order of the assumed mean
generation interval of the variant: 4 days at the top, 5 days in the middle,
and 6 days on the bottom.}

Done.

\revtext{(page 13, lines 282–285) While contact tracing and household studies are
susceptible to a number of biases, they are likely to be a far more reliable
source of information about the transmissibility and generation intervals
of variants than population-level analyses of incidence data.}

Thank you for pointing this out. We have re-written this part:

``Although these estimates are more consistent with the attack rate analysis \citep{ukinvest},
we do not claim that they are necessarily more accurate---while individual-level data from contact tracing can provide a more reliable source of information about the transmissibility and time scale of transmission in some cases, they can be also biased towards particular types of contacts that are easily identified such as household contacts, which could also affect the estimate of $\rho$.''

\revtext{(page 15, References) Curly brackets might need to be added to some of
the BibTeX entries to enforce correct capitalization (e.g., Ref. 9).}

Thank you.

\swp{Ref 9 is ``Infectious diseases of humans: dynamics and control.'' Do we capitalize this or no?}

\rev{Reviewer \#3}

\revtext{In this manuscript, Park and colleagues present a thoughtful analysis of how assumptions about the generation interval distribution for wild-type and variant SARS-CoV-2 can bias estimates of the relative growth rate ("speed") and relative reproduction number ("strength") of the variant. The analysis is mathematically sound and offers a thorough investigation of an important concept that has been the subject of much discussion in the epidemiological community but, to my knowledge, hasn't been given the formal treatment it deserves until now. My comments are mostly minor and intended to make the findings more accessible and transferable to epidemiological practice.}

Thank you.

\revtext{Abstract: 37: On first reading, the distinction between 'speed-like' and 'strength-like' interventions wasn't clear to me, nor why contact tracing should target r but not R, and why social distancing should target R but not r. Could you use more recognizable terms here to highlight how these interventions differ from one another? (is the key difference that contact tracing reduces the number of infectious individuals and thus the probability that a susceptible person comes into contact with an infectious person, while social distancing reduces the per-contact chance of transmission? Or something else?) This comes up again in line 57 of the Introduction.}

Thank you for pointing this out. We have removed 'speed-like' and 'strength-like' interventions from the abstract as these ideas can be confusing without further explanation (which is difficult given word limit). Instead, we have tried to make these ideas clearer throughout text, including in the introduction. We have also revised the abstract according to reviewer 1's comments.

\revtext{166: I think it would be worth stating the range of Rwt and/or Rvar that yields this range for delta (I think it's something like 0.49-1.27 for the wild type, 0.78-2.04 for the variant, based on Figure 2 - but would be helpful if the reader didn't have to derive). Is the varying R what you're referring to when you speak of "underlying epidemiological conditions" in 166-167?}

Thank you for this suggestion.

\revtext{184: I think that "inference data" should be "incidence data"}

Done.

\revtext{Figure 4: I found this figure quite difficult to parse and not especially informative. Are the vertical bars the means of these distributions? If so, they should be labeled and referenced as such in the caption. Also in the figure 4 caption: the term ``generation intervals'' (plural) struck me as a bit strange - I think the singular would make more sense and be more consistent with what's used in the literature. Also, it would be helpful to put theta and phi in the vertical axis labels for E and F (e.g., "Strength (theta)" and "Speed (phi; 1/days)"). Unless this is observed speed and strength, meaning Rvar and rvar? In E, I'm also confused why the strength of the "constant-strength" line is increasing - shouldn't it be flat? Or what have I missed? I think some more details to refer back to the variables in the text might be helpful - for example, I think that (F) is the "Pre-intervention (black) and post-intervention speed of variant transmission (rvar) conditional on the mean generation interval of the variant (Gvar, colored)."}

\revtext{Really what I would like to see here (Figure 4) is some simulated epidemic curves and perhaps final epidemic sizes - I think this would go much further to communicate the intuition behind how the speed-based and strength-based interventions differ than what's currently presented. One of the main conclusions of the paper seems to be that speed-based interventions could be more effective than strength-based interventions when the mean generation interval is long; presenting such simulations would be a good way to drive this point home and to give intuition for just how different speed-based vs strength-based interventions actually are.}

\revtext{264: "epidemiological dynamics" is used again here - what specifically do you mean?}

\revtext{265: "can also lead to different conclusions about the effectiveness of interventions" - what sort of different conclusions? In which circumstances?}

\revtext{271: "If new variants has a longer infectious period" (typo)}

Done.

\revtext{Missing equation reference in the Supplement.}

\bibliography{newvariant_abbv.bib}

\end{document}

### 2021-07-07 newvariant.tex
updating

Inferring the relative strength (i.e., the ratio of reproduction numbers, $\Rv/\Rw$) and relative speed (i.e., the difference between growth rates, $\rv-\rw$) of new SARS-CoV-2 variants compared to their wild types is critical to predicting and controlling the course of the current pandemic.
Multiple studies have estimated the relative strength of new variants from the observed relative speed, but they typically neglect the possibility that the new variants have different generation intervals (i.e., time between infection and transmission), which determines the relationship between relative strength and speed.
Notably, the increasingly predominant B.1.1.7 variant may have a longer infectious period (and therefore, a longer generation interval) than prior dominant lineages.
Here, we explore how differences in generation intervals between a new variant and the wild type affect the relationship between relative strength and speed.
We use simulations to show how neglecting these differences can lead to biases in estimates of relative strength in practice and to illustrate how such biases can be assessed.
Finally, we discuss implications for control: if new variants have longer generation intervals then speed-like interventions such as contact tracing become more effective, whereas strength-like interventions such as social distancing become less effective.

@@

Analyses of new variants have typically characterized \emph{relative} strength (i.e., the ratio of reproduction numbers $\Rv/\Rw$) and speed (i.e., the difference between growth rates $\rv-\rw$) of the variants.
Many studies have tried to estimate the relative strength of variants from the observed relative speed \citep{davies2021estimated, leung2021early, volz2021transmission,zhao2021}.
Some studies have instead assumed a value of the relative strength and tried to predict its relative speed \citep{davies2021estimated,di2021impact}.
While both approaches are reasonable, holding different quantities constant (i.e., strength or speed) can lead to different conclusions about the spread of the disease and its control because two quantities are linked by the generation-interval distribution \citep{doi:10.1098/rspb.2020.1556}.
For example, when epidemic strength is fixed, assuming longer generation intervals lead to a slower epidemic (lower $r$), making the epidemic look easier to control with speed-like interventions such as contact tracing.
When epidemic speed is fixed, assuming longer generation intervals lead to a stronger epidemic (higher $\RR$), making the epidemic look harder to control with strength-like interventions such as social distancing.
Therefore, it is critical to evaluate which assumption is more appropriate for inferring transmission advantages of new variants.

@@

Neglecting the (relatively slow) rate of new mutations, the current incidence of infection $i_x(t)$ caused by each strain $x$---either the wild type (``wt'') or the variant (``var'')---can be expressed in terms of their previous incidence $i_x(t-\tau)$ and the rate at which secondary infections are generated at time $t$ by individuals infected $\tau$ time units ago $K_x(t, \tau)$:
\begin{equation}
i_x(t) = \int_0^\infty i_x(t-\tau) K_x(t, \tau) \dtau.
\end{equation}
This framework provides a flexible way of modeling disease dynamics and generalizes compartmental model, such as the SEIR model \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, champredon2018equivalence}.

@@

Under strength-like changes that reduce the hazard of transmission by a constant amount (e.g., susceptible depletion and decreased transmission rates), the instantaneous generation-interval distribution does not across change time \citep{fraser2007estimating}---
in this case, the instantaneous generation-interval distribution is also often referred to as the intrinsic generation-interval distribution \citep{champredon2015intrinsic,champredon2018two,gostic2020practical,park2020time}.
For example, the standard SEIR model can be equivalently expressed as a renewal equation with time-invariant $g(\tau)$ as shown in \citep{champredon2018equivalence}.
The instantaneous generation-interval distribution is also different from the realized generation-interval distribution, which measures time between actual infection events (see \citep{champredon2015intrinsic}).
Previous studied have noted, in many contexts, that the realized generation intervals can contract due to susceptible depletion---a special case of strength-like changes---due to decreases in transmission even though the instantaneous generation-interval distribution remains unchanged.
In the context of SARS-CoV-2, assuming a time-invariant instantaneous generation-interval distribution is often appropriate given that the dynamics of its spread have been primarily driven by strength-like interventions, especially social distancing measures and vaccination.

Over a short period of time, we can assume that epidemiological conditions remain roughly constant: $\RR_x(t) \approx \RR_x$ and $g_x(t, \tau) \approx g_x(\tau)$.
Then, the incidence of each strain grows (or decays) exponentially at rate $r_x$, satisfying the Euler-Lotka equation \citep{wallinga2007generation}:
\begin{equation}
\frac{1}{\RR_x} = \int_0^\infty \exp(- r_x \tau) g_x(\tau) \dtau.
\end{equation}
We can approximate this $r$--$\RR$ relationship by assuming that the generation-interval distribution is Gamma-distributed, and summarizing it using the mean generation interval $\bar{G}_x$ and the squared coefficient of variation $\kappa_x$ \citep{park2019practical}:
\begin{equation}
\RR_x \approx (1 + \kappa_x r_x \bar{G}_x)^{1/\kappa_x}.
\end{equation}
% The Gamma assumption includes as a special case models that assume exponentially distributed generation intervals (when $\kappa=1$), corresponding to the SIR model \citep{anderson1991infectious}; 
Various Gamma assumptions have been widely used in epidemic modeling, including for models of SARS-CoV-2 \citep{doi:10.1098/rsif.2020.0144}.
We use this framework to investigate how inferences about strength and speed of the variant depend on our assumptions about the underlying generation-interval distributions.
For simplicity we neglect differences in the squared coefficient of variation and assume $\kappa_{\mathrm{wt}} = \kappa_{\mathrm{var}} = \kappa$ and focus on differences in the mean generation intervals.

@@

While epidemic speed $r_x$ can often be estimated directly from incidence of infection during the exponential growth period \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating},
studies of new SARS-CoV-2 variants have mostly focused on characterizing changes in the \emph{proportion} of a new variant \citep{switzerland2021variant, davies2021estimated, di2021impact, graham2021changes, leung2021early, volz2021transmission,zhao2021}.
Focusing on proportions has the advantage, because changes in proportions are less sensitive to changes in testing and to other transient effects that would affect variants and wild type viruses similarly.
When incidence is changing exponentially ($i_x(t) = i_x(t_0) \exp(r_x t)$), the proportion of the new variant $p(t)$ follows a logistic growth curve:
\begin{align}
p(t) &= \frac{\iv(t_0) \exp(\rv t)}{\iw(t_0) \exp(\rw t) + \iv(t_0) \exp(\rv t)},
\\ &= \frac{1}{1 + \left(\iw(t_0)/\iv(t_0)\right) \exp(-\delta t)},
\end{align}
where the logistic growth rate $\delta = \rv - \rw$ corresponds to the relative speed of the epidemic.

@@

Based on the observed relative growth rate of B.1.1.7 in the UK, we start by taking the relative speed of the variant to be $\delta = 0.1\pday$  \citep{davies2021estimated}; the mean generation interval of the wild type to be $\Gw = 5\days$ \citep{ferretti2020quantifying}; and the squared coefficient of variation of generation intervals to be $\kappa=0.2$ \citep{ferretti2020quantifying} for both the variant and the wild type.
% As noted above, we assume that the variant and the wild type have equal $\kappa$ throughout, and only consider differences in the mean.
We evaluate the estimates of relative strength $\rho$ across a wide range of $\kappa$ from 0 (fixed-length generation intervals) to 1 (exponential distribution).
To further explore how inference depends on underlying epidemiological conditions, we consider five scenarios: (1) $\rw < \rv < 0$, (2) $\rw < \rv = 0$, (3) $\rw < 0 < \rv$, (4) $0 = \rw < \rv$, and (5) $0 < \rw < \rv$.
We consider the spread of B.1.1.7 as an example but our qualitative conclusions should hold for other VoCs.

@@

We thus investigate how $\delta$ is expected to change with \Rw if $\rho$ remains constant, and how this expectation changes with the ratio of the generation intervals. 
Once again, we rely on the Gamma assumption \citep{park2019practical} to find the relative speed $\delta$ given the mean generation interval of the variant $\Gv$ and the wild type $\Gw$:
\begin{equation}
\delta = \frac{(\rho \Rw)^{\kappa} - 1}{\kappa \Gv} - \frac{\Rw^{\kappa} - 1}{\kappa \Gw}.
\end{equation}
As our baseline scenario we assume $\rho = 1.61$, which is the value we obtain for $\delta=0.1\pday$ \citep{davies2021estimated}, $\rw=0\pday$, $\Gw = \Gv = 5\,\textrm{days}$, and $\kappa = 1/5$ \citep{ferretti2020quantifying}.
We evaluate $\delta$ across five scenarios as before: (1) $\Rw < \Rv < 1$, (2) $\Rw < \Rv = 1$, (3) $\Rw < 1 < \Rv$, (4) $1 = \Rw < \Rv$, and (5) $1 < \Rw < \Rv$.

@@

\fref{relspeed} also shows that when $\rho$ is fixed relative speed depends on underlying epidemiological conditions. 
For example, even when there are no differences in the generation-interval distributions ($\Gv=\Gw$, in this case), the relative speed $\delta$ can range between 0.08--0.11 when $\kappa=0.2$ and 0.06--0.14 when $\kappa=0.5$.
Therefore, characterizing the spread of variants assuming constant relative speed (by fitting a standard logistic growth curve; e.g., \citep{campbell2021increased,switzerland2021variant}) should be used with care.

@@

As new variants of SARS-CoV-2, such as B.1.1.7, are spreading and becoming dominant in many countries, it is clear that the variants are more transmissible (have higher strength) than the wild type \citep{switzerland2021variant, davies2021estimated, di2021impact, graham2021changes, leung2021early, volz2021transmission,zhao2021}.
While our analysis supports estimates of a higher strength of the new variant across a wide range of assumptions, it also shows that uncertainty in generation-interval distributions must be taken into account to obtain accurate estimates of the relative strength of variants.
If new variants has a longer infectious period \citep{kissler2021densely}, and therefore extended generation-interval distributions, current estimates may underestimate the true relative strength.
% If new variants escape vaccinal or natural immunity to some extent \citep{Kupferschmidt329}, their relative strength may increase further.

>>>>

Inferring the relative strength (i.e., the ratio of reproduction numbers, $\Rv/\Rw$) and relative speed (i.e., the difference between growth rates, $\rv-\rw$) of new SARS-CoV-2 variants compared to their wild types is critical to predicting and controlling the course of the current pandemic.
Although the role of the generation interval (i.e., time between infection and transmission) in linking epidemic strength and speed has been previously well-established, many analyses of new SARS-CoV-2 variants overlook uncertainties in the generation-interval distributions as well as the possibility that their inferences of epidemic dynamics can depend on whether relative strength or speed are held constant.
Here, we explore how differences in the generation-interval distributions affect the inference of relative strength and speed in light of recent evidence that the B.1.1.7 variant may have a longer infectious period (and therefore, a longer generation interval) than prior dominant lineages.
We use simulations to show how neglecting these differences can lead to biases in estimates of relative strength in practice and to illustrate how such biases can be assessed.
We also assess which of the two assumptions (i.e., constant relative strength or speed) are more appropriate and further discuss implications for control.

@@

Analyses of new variants have typically characterized \emph{relative} strength (i.e., the ratio of reproduction numbers $\Rv/\Rw$) and speed (i.e., the difference between growth rates $\rv-\rw$) of the variants.
Many studies have tried to estimate the relative strength of variants from the observed relative speed \citep{davies2021estimated, leung2021early, volz2021transmission,zhao2021}.
Some studies have instead assumed a value of the relative strength and tried to predict its relative speed \citep{davies2021estimated,di2021impact}.
While both approaches are reasonable, holding different quantities constant (i.e., strength or speed) can lead to different conclusions about the spread of the disease and its control because two quantities are linked by the generation-interval distribution \citep{doi:10.1098/rspb.2020.1556}.
For example, when epidemic strength is fixed, assuming longer generation intervals lead to a slower epidemic (lower $r$), making the epidemic look easier to control with speed-like interventions (e.g., contact tracing), which depend on hazard of isolation and therefore affects late transmission disproportionately.
When epidemic speed is fixed, assuming longer generation intervals lead to a stronger epidemic (higher $\RR$), making the epidemic look harder to control with strength-like interventions (e.g., social distancing), which reduce the hazard of infectious contact and therefore affecting transmission by an equal amount regardless of age of infection.
Therefore, it is critical to evaluate which assumption is more appropriate for inferring transmission advantages of new variants.

@@

Neglecting the (relatively slow) rate of new mutations, the current incidence of infection $i_x(t)$ caused by each strain $x$---either the wild type (``wt'') or the variant (``var'')---can be expressed in terms of their previous incidence $i_x(t-\tau)$ and the rate at which secondary infections are generated at time $t$ by individuals infected $\tau$ time units ago $K_x(t, \tau)$:
\begin{equation}
i_x(t) = \int_0^\infty i_x(t-\tau) K_x(t, \tau) \dtau.
\end{equation}
This framework provides a flexible way of modeling disease dynamics and generalizes compartmental model, such as the SEIR model \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral,breda2012formulation, champredon2018equivalence}.

@@

Under strength-like changes that reduce the hazard of transmission by a constant amount (e.g., susceptible depletion and decreased transmission rates), the instantaneous generation-interval distribution does not across change time \citep{fraser2007estimating}.
In this case, the instantaneous generation-interval distribution is also often referred to as the intrinsic generation-interval distribution \citep{champredon2015intrinsic,champredon2018two,gostic2020practical,park2020time}---
for example, the standard SEIR model can be equivalently expressed as a renewal equation with time-invariant intrinsic generation-interval distribution $g(\tau)$ as shown in \citep{champredon2018equivalence}.
The instantaneous generation-interval distribution is also different from the realized generation-interval distribution, which measures time between actual infection events (see \citep{champredon2015intrinsic}).
Previous studied have noted, in many contexts, that the realized generation intervals can contract due to susceptible depletion---a special case of strength-like changes---due to decreases in transmission \citep{kenah2008generation,nishiura2010time,champredon2015intrinsic}, even though the instantaneous generation-interval distribution remains unchanged.
While the instantaneous generation-interval distribution can change under speed-like changes that disproportionately affect late transmissions (see \cite{fraser2007estimating} and Section 6 for discussion), assuming a time-invariant instantaneous generation-interval distribution is often appropriate in the context of SARS-CoV-2, given that the dynamics of its spread have been primarily driven by strength-like interventions, such as social distancing measures \citep{flaxman2020Rt} and vaccination \citep{moore2021vaccination}.
Therefore, we neglect changes in the intrinsic generation-interval distribution over time for now and return to this idea in Section 6.

Over a short period of time, we can assume that epidemiological conditions remain roughly constant: $\RR_x(t) \approx \RR_x$ and $g_x(t, \tau) \approx g_x(\tau)$.
Here, we use the term ``epidemiological conditions'' to broadly refer to all factors that affect transmission---mathematically, they are captured by the kernel $K(t, \tau)$.
In the context of SARS-CoV-2, we are essentially assuming that the changes in susceptible pool, behavior, and contact rates are negligible over a short period of time.
Then, the incidence of each strain grows (or decays) exponentially at rate $r_x$, satisfying the Euler-Lotka equation \citep{wallinga2007generation}:
\begin{equation}
\frac{1}{\RR_x} = \int_0^\infty \exp(- r_x \tau) g_x(\tau) \dtau.
\end{equation}
We can approximate this $r$--$\RR$ relationship by assuming that the generation-interval distribution is Gamma-distributed, and summarizing it using the mean generation interval $\bar{G}_x$ and the squared coefficient of variation $\kappa_x$ \citep{park2019practical}:
\begin{equation}
\RR_x \approx (1 + \kappa_x r_x \bar{G}_x)^{1/\kappa_x}.
\end{equation}
% The Gamma assumption includes as a special case models that assume exponentially distributed generation intervals (when $\kappa=1$), corresponding to the SIR model \citep{anderson1991infectious}; 
Various Gamma assumptions have been widely used in epidemic modeling, including for models of SARS-CoV-2 \citep{doi:10.1098/rsif.2020.0144}.
We use this framework to investigate how inferences about strength and speed of the variant depend on our assumptions about the underlying generation-interval distributions.
For simplicity we neglect differences in the squared coefficient of variation and assume $\kappa_{\mathrm{wt}} = \kappa_{\mathrm{var}} = \kappa$.
While the exact shape of the distribution matters, we choose to focus on differences in the mean generation intervals to provide a qualitative understanding of how differences in generation-interval distributions between the wild type and the variant affects the relationship between relative strength and speed.
In general, we expect  differences in the squared coefficient of variation to have qualitatively similar effects as differences in the mean.
For example, a wider distribution allows for more early transmission, which in turn reduces epidemic strength given speed---this effect is qualitatively similar to that of a distribution with a shorter mean.

@@

While epidemic speed $r_x$ can often be estimated directly from incidence of infection during the exponential growth period \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating},
studies of new SARS-CoV-2 variants have mostly focused on characterizing changes in the \emph{proportion} of a new variant \citep{switzerland2021variant, davies2021estimated, di2021impact, graham2021changes, leung2021early, volz2021transmission,zhao2021}.
Focusing on proportions has the advantage, because changes in proportions are less sensitive to changes in testing and to other transient effects that would affect variants and wild type viruses similarly.
When incidence is changing exponentially ($i_x(t) = i_x(t_0) \exp(r_x t)$), the proportion of the new variant $p(t)$ follows a logistic growth curve \citep{switzerland2021variant,davies2021estimated}:
\begin{align}
p(t) &= \frac{\iv(t_0) \exp(\rv t)}{\iw(t_0) \exp(\rw t) + \iv(t_0) \exp(\rv t)},
\\ &= \frac{1}{1 + \left(\iw(t_0)/\iv(t_0)\right) \exp(-\delta t)},
\end{align}
where the logistic growth rate $\delta = \rv - \rw$ (i.e., the slope of the linear regression at the logit scale) corresponds to the relative speed of the epidemic.
When more than two strains are co-circulating, the growth rate at a logit scale is expected to change over time and does not correspond to the relative speed even when all strains exhibit exponential changes (see Supplementary Text);
instead, as shown in \citep{campbell2021increased}, a multinomial logistic regression model is required to infer relative speed.
Since we are interested in the relationship between the true relative speed and strength, rather than the inference of relative speed from proportion data, our results are insensitive to the number of strains circulating in the population.

@@

Based on the observed relative growth rate of B.1.1.7 in the UK, we start by taking the relative speed of the variant to be $\delta = 0.1\pday$  \citep{davies2021estimated}; the mean generation interval of the wild type to be $\Gw = 5\days$ \citep{ferretti2020quantifying}; and the squared coefficient of variation of generation intervals to be $\kappa=0.2$ \citep{ferretti2020quantifying} for both the variant and the wild type.
% As noted above, we assume that the variant and the wild type have equal $\kappa$ throughout, and only consider differences in the mean.
We evaluate the estimates of relative strength $\rho$ across a wide range of $\kappa$ from 0 (fixed-length generation intervals) to 1 (exponential distribution).
To further explore how inference depends on underlying epidemiological conditions, we consider five scenarios: (1) $\rw < \rv < 0$, (2) $\rw < \rv = 0$, (3) $\rw < 0 < \rv$, (4) $0 = \rw < \rv$, and (5) $0 < \rw < \rv$.
In all of these cases, we always assume $\rw < \rv$ to reflect the observed relative speed of $\delta = 0.1\pday$.
We consider the spread of B.1.1.7 as an example but our qualitative conclusions should hold for other VoCs.

@@

We thus investigate how $\delta$ is expected to change with \Rw if $\rho$ remains constant, and how this expectation changes with the ratio of the generation intervals. 
Once again, we rely on the Gamma assumption \citep{park2019practical} to find the relative speed $\delta$ given the mean generation interval of the variant $\Gv$ and the wild type $\Gw$:
\begin{equation}
\delta = \frac{(\rho \Rw)^{\kappa} - 1}{\kappa \Gv} - \frac{\Rw^{\kappa} - 1}{\kappa \Gw}.
\end{equation}
As our baseline scenario we assume $\rho = 1.61$ (and therefore, we always have $\Rw < \Rv$), which is the value we obtain for $\delta=0.1\pday$ \citep{davies2021estimated}, $\rw=0\pday$, $\Gw = \Gv = 5\,\textrm{days}$, and $\kappa = 1/5$ \citep{ferretti2020quantifying}.
We evaluate $\delta$ across five scenarios as before: (1) $\Rw < \Rv < 1$, (2) $\Rw < \Rv = 1$, (3) $\Rw < 1 < \Rv$, (4) $1 = \Rw < \Rv$, and (5) $1 < \Rw < \Rv$.

@@

\fref{relspeed} also shows that when $\rho$ is fixed relative speed depends on underlying epidemiological conditions---that is, whether the incidence of wild type and the variant are increasing or decreasing. 
For example, even when there are no differences in the generation-interval distributions ($\Gv=\Gw$, in this case), changing $\Rw$ from 0.49 to 1.27 (and thus $\Rv$ from 0.79 to 2.04) yields the relative speed $\delta$ between 0.08--0.11 for $\kappa=0.2$ and 0.06--0.14 for $\kappa=0.5$.
Therefore, characterizing the spread of variants assuming constant relative speed (by fitting a standard logistic growth curve; e.g., \citep{campbell2021increased,switzerland2021variant}) should be used with care.

@@

As new variants of SARS-CoV-2, such as B.1.1.7, are spreading and becoming dominant in many countries, it is clear that the variants are more transmissible (have higher strength) than the wild type \citep{switzerland2021variant, davies2021estimated, di2021impact, graham2021changes, leung2021early, volz2021transmission,zhao2021}.
While our analysis supports estimates of a higher strength of the new variant across a wide range of assumptions, it also shows that uncertainty in generation-interval distributions must be taken into account to obtain accurate estimates of the relative strength of variants.
If new variants have a longer infectious period \citep{kissler2021densely}, and therefore extended generation-interval distributions, current estimates may underestimate the true relative strength.
% If new variants escape vaccinal or natural immunity to some extent \citep{Kupferschmidt329}, their relative strength may increase further.
